\ripple: Differentiable Waveforms for Gravitational Wave Data Analysis

### 2022-11-10 src/tex/ms.tex
Wrote some test for the Fisher forecasting

The three major waveform families that have been developed to date are: the Effective-one-body (EOB), numerical relativity (NR) surrogate, and phenomenological inspiral-merger-ringdown (PhenomIMR).
Note that unlike NR surrogate, both the EOB and PhenomIMR waveforms are calibrated to NR waveforms.
EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate whereas PhenomIMR waveforms are constructed with simple closed-form expressions.
PhenomIMR waveforms are therefore ideally suited for AD, espeically using \jax. 
In this paper we focus mainly on the aligned spin circular orbit model IMRPhenomD~\citep{Husa:2015iqa, 2015jqa}.

@@

Here, we will illustrate how three core tasks in GW science can be substantially improved through the use of differentiable waveforms.
In the paper we will primarily look at toy examples, leaving a more careful analysis to future work. 
The three tasks discussed here cover a wide range of GW science, starting with waveform development all the way to parameter estimation.

@@

\subsection{Effectualness Calculations}
\label{sec:effectualness}

@@

>>>>

The three major waveform families that have been developed to date are: the Effective-one-body (EOB), numerical relativity (NR) surrogate, and phenomenological inspiral-merger-ringdown (PhenomIMR).
Note that unlike NR surrogate, both the EOB and PhenomIMR waveforms are calibrated to NR waveforms.
EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate whereas PhenomIMR waveforms are constructed with simple closed-form expressions.
PhenomIMR waveforms are therefore ideally suited for AD, espeically using \jax. 
In this paper we focus mainly on the aligned spin circular orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.

@@

Here, we will illustrate how three core tasks in GW science can be substantially improved through the use of differentiable waveforms.
In the paper we will primarily look at toy examples, leaving a more careful analyses to future work. 
The three tasks discussed here cover a wide range of GW science, starting with waveform development all the way to forecasting and parameter estimation.

@@

\te{Kelvin should add stuff here}

\subsection{Fisher Forecasting}
\label{subsec:hmc}

Forecasting the sensitivity of future experiments is an essential task in GW science.
Due to its theoretical simplicity and its evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters can be measured.
The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
Although in practice this assumption is often violated in real GW datasets, the results obtained using a Fisher analysis can provide quick and useful diagnotistics in evaluating sensitivities for a variety of models and detector configurations.

Computing the Fisher matrix requires one to evaluate derivatives of the likelihhood which in turn involves derivatives of the waveform model and detector projection functions.
AD is therefore perfectly suited for computing Fisher matrices accurately and efficiently. 
Forecasting with Fisher matrices for third generation detectors has already been extensively explored in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
Here we purely want to illustrate the simplicitly and speed of AD for forecasting rather than providing new physics insights.
We therefore consider a simple, three detector setup corresponding to the two LIGO detectors in addition to Virgo.

The Fisher information matrix for a single detector is typically given by 
\begin{equation}
    F^{k}_{ij} = (\partial_i h_0 | \partial_j h_0) \, ,
\end{equation}
where $k$ indicates the detector, $\partial_i = \partial/\partial \theta_i$, and $h_0$ is the strain measured by the detector which is given by,
\begin{equation}
    h_0(\theta) = F_+(\lambda) h_{+}(\phi) + F_\times(\lambda) h_{\times}(\phi) \, .
\end{equation}
Here we have separated out the intrinsic ($\phi$) and extrinsic ($\lambda$) variables.
Since we are considering a three detector setup we simply add the Fisher matrices from the individual detectors to get the combined Fisher matrix:
\begin{equation}
    F_{ij} =  F^{\mathrm{Hanford}}_{ij} + F^{\mathrm{Livingston}}_{ij} + F^{\mathrm{Virgo}}_{ij}   \, .
\end{equation}
Finally, we invert the Fisher matrix to calculate the covariance matrix.

To illustrate the computational speed of computing Fisher matrices with AD, we consider a population of binaries and compute the sky localtization error following Eq.~(28) in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
Since the Fisher matrix approach is known to have both theoretical issues as well as numerical instabilities for low signal-to-noise events, we restrict our population to only nearby systems.
A full list of the distributions used to generate the various parameters in our population are given in Tab...

The distribution of sky localization errors from a a population of $10^3$ binaries can be seen in Fig... 
We have verified that our errors agree with a separate dedicated Fisher forecasting code~\citep{Borhanian:2020ypi} to within X percent.
This demonstrates that AD can be used to accurately produce population level forecasts.

Moreover, each error calculation (including computing the Fisher matrices for each detector and the inversion procress) is substantially faster.
In particular, we find that after compilation each error calculation takes approximately one second on a single computing core.
This can be compared to \texttt{GWbench}~\citep{Borhanian:2020ypi} where each Fisher calculation takes $\mathcal{O}$(minutes) for a similar detector setup and frequency grid.
The factor of almost 60 speed up is despite the fact that a single core evaluation of the \ripple waveform is significantly slower that the \lalsuite call.
As discussed above, performance can be further improved by utilising harware acceleration such as parallel GPU processing.
AD therefore represents a fast and accurate way of performing population level analyses, and should be utilized for testing the capabilities of next generation detectors.

@@

\section{Discussion}
\label{subsec:discussion}
